Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-21 paper.tex
Pass through GEL section

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step 
process to limit memory usage and ensure reliability. We 
first converted the 28 VCF files (3.51 TiB) for chromosome 20
into the intermediate columnar format (ICF). This task was 
split into 3899 partitions, and distributed using the Genomics England
HPC cluster. It required about 50 days of CPU, with an elapsed
wall-clock time of about 2 hours. The ICF representation uses a total
of XTiB over XX files and YY directories. We then encoded this intermediate
columnar format to Zarr using 8 cores and XX GB of RAM in about 12 hours.
This produced a dataset with 43 arrays, consuming a 
total of Y GiB of storage over X directories and 
Y files. This is about a 4.8X reduction over the original VCF. 
The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
default settings. Each field is stored independently 
as a Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in 
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes.
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
/call\_AD & int16 & 658.4G & 26.0 & 25.35\% \\
/call\_GQ & int16 & 654.5G & 13.0 & 25.20\% \\
/call\_DP & int16 & 570.0G & 15.0 & 21.95\% \\
/call\_DPF & int16 & 447.1G & 20.0 & 17.22\% \\
/call\_PL & int16 & 162.6G & 160.0 & 6.26\% \\
/call\_GQX & int16 & 41.0G & 210.0 & 1.58\% \\
/call\_FT & string & 25.0G & 1400.0 & 0.96\% \\
/call\_genotype & int8 & 21.5G & 410.0 & 0.83\% \\
/call\_genotype\_mask & bool & 12.8G & 680.0 & 0.49\% \\
/call\_genotype\_phased & bool & 2.4G & 1900.0 & 0.09\% \\
/call\_PS & int8 & 383.4M & 12000.0 & 0.01\% \\
% /call\_ADF & int8 & 383.4M & 12000.0 & 0.01\% \\
% /call\_ADR & int8 & 383.4M & 12000.0 & 0.01\% \\
/variant\_position & int32 & 111.6M & 2.0 & \\
/variant\_quality & float32 & 87.4M & 2.6 & \\
/variant\_allele & string & 69.3M & 13.0 & \\
% /variant\_OLD\_MULTIALLELIC & object & 55.88M & 8.2 & 0.00\% \\
/variant\_AN & int32 & 47.3M & 4.8 & \\
% /variant\_AC & int32 & 47M & 4.9 & 0.00\% \\
% /variant\_AC\_Het & int32 & 46.62M & 4.9 & 0.00\% \\
% /variant\_ABratio & float32 & 20.23M & 11.0 & 0.00\% \\
% /variant\_MendelSite & float32 & 20.09M & 11.0 & 0.00\% \\
% /variant\_medianGQ & float32 & 19.16M & 12.0 & 0.00\% \\
% /variant\_completeGTRatio & float32 & 14.98M & 15.0 & 0.00\% \\
% /variant\_phwe\_eur & float32 & 14.46M & 16.0 & 0.00\% \\
% /variant\_phwe\_afr & float32 & 13.37M & 17.0 & 0.00\% \\
% /variant\_medianDepthNonMiss & float32 & 12.17M & 19.0 & 0.00\% \\
% /variant\_medianDepthAll & float32 & 11.99M & 19.0 & 0.00\% \\
% /variant\_phwe\_sas & float32 & 11.59M & 20.0 & 0.00\% \\
% /variant\_AC\_Hom & int32 & 11.12M & 21.0 & 0.00\% \\
% /variant\_missingness & float32 & 9.3M & 25.0 & 0.00\% \\
% /variant\_phwe\_eas & float32 & 7.12M & 32.0 & 0.00\% \\
/variant\_filter & bool & 6.4M & 570.0 & \\
% /variant\_AC\_Hemi & int32 & 849.43 KiB & 280.0 & 0.00\% \\
% /variant\_composite\_filter & bool & 433.38 KiB & 130.0 & 0.00\% \\
% /variant\_OLD\_CLUMPED & object & 384.18 KiB & 1200.0 & 0.00\% \\
% /variant\_id & object & 298.76 KiB & 1600.0 & 0.00\% \\
% /variant\_phwe\_amr & float32 & 269.52 KiB & 870.0 & 0.00\% \\
% /variant\_id\_mask & bool & 269.46 KiB & 220.0 & 0.00\% \\
% /sample\_id & object & 268.12 KiB & 2.3 & 0.00\% \\
% /variant\_contig & int16 & 257.77 KiB & 450.0 & 0.00\% \\
% /contig\_length & int64 & 3.79 KiB & 5.3 & 0.00\% \\
% /contig\_id & object & 2.3 KiB & 8.8 & 0.00\% \\
% /filter\_id & object & 808 bytes & 0.6 & 0.00\% \\
%%%% OLD for chr20
% call\_AD &  int16 & 179.7G & 26 & 24.0\\
% call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
% call\_DP &  int16 & 141.8G & 16 & 18.9 \\
% call\_DPF& int16  & 115.1G & 20 & 15.3\\
% call\_FT &  string & 58.5G & 160 & 7.8 \\
% call\_PL &  int16 & 51.4G & 140 & 6.9 \\
% call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
% call\_genotype & int8 & 6.1G & 380 & 0.8 \\
% call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
% call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
% call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
% variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
% variant\_allele & string & 21.5M & 11 \\
% variant\_position & int32 & 12.7M & 4.7 \\
% variant\_ABratio & float32 & 9.6M & 6.3 \\
% variant\_AN & int32 & 9.6M & 6.3 \\
% variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab-genomics-england-data} shows that the dataset storage
size is dominated by a few columns with the top four
(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for over 
80\% of the total. These fields are much less compressible
than genotype data (which uses $<1\%$ of the total space here)
because of their inherent noisiness~\citep{lin2020sparse}.

[TODO add in benchmarks discussed in issue. Extract POS, run af-dist, 
do a filtering query. Note problems with orchestration when
dealing with multiple VCF chunks. 
Talk about mask-oriented computing there at the
end, pointing out the cost of these multiple copies.]

% By default the \texttt{bio2zarr} conversion is lossless, and integers
% of the narrowest type required to represent all of the data are
% used. In this case, this has resulted in 16 bit integers being used 
% for these top four fields. Examining the distribution of the depth
% fields, (call\_AD, call\_DP, call\_DPF), we can 
% we can see that a tiny fraction of the values stored
% are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
% Fig SX for histograms).
% One might then reasonably truncate these fields to 8 bit integers.
% After converting to 8 bit, call\_AD is reduced by 
% x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
% and call\_DPF by x\% (XX GiB). 
% Similarly, the vast majority of call\_GQ values (xx\%) are less than 
% 128 (corresponding to a conditional probability that the genotype call is 
% wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
% be truncated. Here, this results in a reduction of x\% (XX GiB).
% Combined, truncating these top four fields to 8 bits reduces the 
% overall storage of the dataset to XXGiB, corresponding to a 
% X-fold compression of the VCF.

% and can reasonably be truncated (a 
% much information would be lost if we used 8 bit
% Substantial
% reductions in space can be achieved, however, if we examine the 
% distributions of these fields and truncate some outliers.
% Figure~SX shows the histogram of the XX call\_AD values, and we can
% see that < 0.0X\% of the values are $>127$. Truncating and 
% re-encoding the call\_AD array as an int8 results in a total storage of 
% XX. A substantial advantage of Zarr is that such per-field optimisations
% can be made after conversion, without affecting the rest of the dataset.
% As discussed in Methods, judicious use of additional filters
% and lossy quantizing can result substantial space savings.

% Calculating the histogram of per call values at this 
% scale is not a trivial task.
% Taking advantage of the rich software ecosystem growing around Zarr
% (see Discussion) we used xarray~\citep{hoyer2017xarray}
% and Dask~\citep{rocklin2015dask} to transparently distribute the 
% computations over a cluster of X workers. 
% [ADD DETAILS]
% The elapsed time was X minutes, and requires X lines of Python in a 
% Jupyter notebook~\citep{kluyver2016jupyter}.

% One of the key uses of the aggV2 dataset is to enable simple queries 
% of specific regions. For example, in [A PAPER] they [DID SOMETHING].
% We replicated this by running the appropriate bcftools query
% on the VCF chunk (note that the user is responsible for finding 
% the correct chunk file here, limiting the utility of genomic 
% indexes). This took XXX hours.
% The same query implemented with the Zarr-python 
% API required XXX seconds to produce identical output.

>>>>

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step 
process to limit memory usage and ensure reliability. We 
first converted the 106 VCF files (12.81 TiB) for chromosome 2
into the intermediate columnar format (ICF). This task was 
split into 14,605 partitions, and distributed using the Genomics England
HPC cluster. The average run-time per partition was 20.7 min.
The ICF representation used a total
of 9.94 TiB over 3,960,177 data storage files. 
We then converted the ICF to Zarr [DETAILS].
This produced a dataset with 44 arrays, consuming a 
total of 2.54 TiB of storage over 6,312,488
chunk files. This is a roughly 5X reduction in total storage space over the original VCF. 
The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
Genomics England aggV2 VCFs on chromosome 2 using \texttt{vcf2zarr}
default settings. Each field is stored independently 
as a Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in 
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes
(omitting values < 0.01\%).
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
/call\_AD & int16 & 658.4G & 26.0 & 25.35\% \\
/call\_GQ & int16 & 654.5G & 13.0 & 25.20\% \\
/call\_DP & int16 & 570.0G & 15.0 & 21.95\% \\
/call\_DPF & int16 & 447.1G & 20.0 & 17.22\% \\
/call\_PL & int16 & 162.6G & 160.0 & 6.26\% \\
/call\_GQX & int16 & 41.0G & 210.0 & 1.58\% \\
/call\_FT & string & 25.0G & 1400.0 & 0.96\% \\
/call\_genotype & int8 & 21.5G & 410.0 & 0.83\% \\
/call\_genotype\_mask & bool & 12.8G & 680.0 & 0.49\% \\
/call\_genotype\_phased & bool & 2.4G & 1900.0 & 0.09\% \\
/call\_PS & int8 & 383.4M & 12000.0 & 0.01\% \\
% /call\_ADF & int8 & 383.4M & 12000.0 & 0.01\% \\
% /call\_ADR & int8 & 383.4M & 12000.0 & 0.01\% \\
/variant\_position & int32 & 111.6M & 2.0 & \\
/variant\_quality & float32 & 87.4M & 2.6 & \\
/variant\_allele & string & 69.3M & 13.0 & \\
% /variant\_OLD\_MULTIALLELIC & object & 55.88M & 8.2 & 0.00\% \\
/variant\_AN & int32 & 47.3M & 4.8 & \\
% /variant\_AC & int32 & 47M & 4.9 & 0.00\% \\
% /variant\_AC\_Het & int32 & 46.62M & 4.9 & 0.00\% \\
% /variant\_ABratio & float32 & 20.23M & 11.0 & 0.00\% \\
% /variant\_MendelSite & float32 & 20.09M & 11.0 & 0.00\% \\
% /variant\_medianGQ & float32 & 19.16M & 12.0 & 0.00\% \\
% /variant\_completeGTRatio & float32 & 14.98M & 15.0 & 0.00\% \\
% /variant\_phwe\_eur & float32 & 14.46M & 16.0 & 0.00\% \\
% /variant\_phwe\_afr & float32 & 13.37M & 17.0 & 0.00\% \\
% /variant\_medianDepthNonMiss & float32 & 12.17M & 19.0 & 0.00\% \\
% /variant\_medianDepthAll & float32 & 11.99M & 19.0 & 0.00\% \\
% /variant\_phwe\_sas & float32 & 11.59M & 20.0 & 0.00\% \\
% /variant\_AC\_Hom & int32 & 11.12M & 21.0 & 0.00\% \\
% /variant\_missingness & float32 & 9.3M & 25.0 & 0.00\% \\
% /variant\_phwe\_eas & float32 & 7.12M & 32.0 & 0.00\% \\
/variant\_filter & bool & 6.4M & 570.0 & \\
% /variant\_AC\_Hemi & int32 & 849.43 KiB & 280.0 & 0.00\% \\
% /variant\_composite\_filter & bool & 433.38 KiB & 130.0 & 0.00\% \\
% /variant\_OLD\_CLUMPED & object & 384.18 KiB & 1200.0 & 0.00\% \\
% /variant\_id & object & 298.76 KiB & 1600.0 & 0.00\% \\
% /variant\_phwe\_amr & float32 & 269.52 KiB & 870.0 & 0.00\% \\
% /variant\_id\_mask & bool & 269.46 KiB & 220.0 & 0.00\% \\
/sample\_id & str & 268.12K & 2.3 & \\
% /variant\_contig & int16 & 257.77 KiB & 450.0 & 0.00\% \\
% /contig\_length & int64 & 3.79 KiB & 5.3 & 0.00\% \\
% /contig\_id & object & 2.3 KiB & 8.8 & 0.00\% \\
% /filter\_id & object & 808 bytes & 0.6 & 0.00\% \\
%%%% OLD for chr20
% call\_AD &  int16 & 179.7G & 26 & 24.0\\
% call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
% call\_DP &  int16 & 141.8G & 16 & 18.9 \\
% call\_DPF& int16  & 115.1G & 20 & 15.3\\
% call\_FT &  string & 58.5G & 160 & 7.8 \\
% call\_PL &  int16 & 51.4G & 140 & 6.9 \\
% call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
% call\_genotype & int8 & 6.1G & 380 & 0.8 \\
% call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
% call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
% call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
% variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
% variant\_allele & string & 21.5M & 11 \\
% variant\_position & int32 & 12.7M & 4.7 \\
% variant\_ABratio & float32 & 9.6M & 6.3 \\
% variant\_AN & int32 & 9.6M & 6.3 \\
% variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab-genomics-england-data} shows that the dataset storage
size is dominated by a few columns with the top four
(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for 
90\% of the total. These fields are much less compressible
than genotype data (which uses $<1\%$ of the total space here)
because of their inherent noisiness~\citep{lin2020sparse}. 
Note that these top four fields are stored as 16 bit integers
because they contain rare outliers that cannot be stored as 
8 bits. While the fields could likely be truncated to have 
a maximum of 127 with minimal loss of information, 
the compression gains from doing so are relatively minor,
and we therefore opt for fully lossless compression here 
for simplicity.

To demonstrate the computational accessibility of Zarr on this 
large human dataset, we performed some illustrative benchmarks.
Firstly, we ran the af-dist calculation (Figs~\ref{fig-whole-matrix-compute}
and~\ref{fig-subset-matrix-compute}) on one 132GiB VCF file covering
positions 58,219,159--60,650,943 (562,640 variants) using \texttt{bcftools
+af-dist} as before. The elapsed time for this operation was 716 min
user+sys time almost exactly equal). We repeated this operation for the 
same genome coordinates in Zarr (using Python code described in previous sections) 
% 716 minutes / 4.25 = 168X
with a total CPU time of 2min 19s and elapsed time of 4min 15s (168X faster).
Second, we extracted the genome position for each variant in this single VCF 
chunk using \texttt{bcftools query} and Python Zarr code as described in 
Fig~\ref{fig-column-extract}. The \texttt{bcftools} command required 
% def parse_time(time_str):
%     minsplit = time_str.split("m")
%     return datetime.timedelta(minutes=int(minsplit[0]),
%         seconds=float(minsplit[1][:-1]))
% def parse_time_output(s):
%     values = {}
%     for line in s.splitlines():
%         split = line.split()
%         values[split[0]] = parse_time(split[1])
%     return values["real"].seconds / 60, (values["user"] + values["sys"]).seconds / 60
% real  85m51.013s
% user  54m39.986s
% sys 0m45.277s
55.42 min total CPU (85.85 min elapsed).
% CPU times: user 846 ms, sys: 1.93 s, total: 2.78 s
% Wall time: 1min 44s
% 85.85 / 1.73 = 49.62 = 50X
The Zarr code required 2.78 sec total CPU and 1.73 min elapsed (50X faster).
Note that the Zarr code here is heavily I/O bound [TODO say something 
about the GEL cluster?]. When we subsequently time the operation of 
just populating an array of the variant positions in specified region,
this is reduced to 1.13 sec, with an elapsed time of 20.9 sec.
[Note: this is crazy that we still have a 19 second I/O wait. The
5989 files for the POS chunks must be in cache already, so the delay
has to be the file system checking cache coherency.]

Finally, to illustrate performance on a common filtering task, we 
created a copy of the VCF chunk which contains only variants 
that pass some common filtering criteria using
\texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
following standard practices [citations?].
This took a total of 689 min CPU time, with the same elapsed time.
In comparison, computing and storing a variant mask (i.e., a boolean value 
for each variant denoting whether it should be considered or not
for analysis) based on the same criteria using Zarr 
required a total of 1.96 min CPU time with an elapsed time 
of  11 min (63X faster).
%689 / 11 = 62.6 = 63X
There is an important distinction here between creating a copy
of the data (an implicit part of VCF based workflows) and creating an
an additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
illustrates, call-level masks are cheap (the standard 
missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
storage) and variant or sample level masks requiring negligible storage.
If downstream software can 
use configurable masks (at variant, sample and call level) 
rather than expecting full copies of the data, major storage savings
and improvements in processing efficiency can be made.
Such a transition from ``mask-oriented'' vs ``copy-oriented'' 
analysis is one of the transformational changes enabled by Zarr.
